taxize - taxonomic search and retrieval in R

### 2013-04-27 inst/doc/taxize_withcode.Rnw
updated the mansucript

\documentclass[letterpaper,superscriptaddress,showkeys,longbibliography]{revtex4-1}
\usepackage[utf8]{inputenc}
\usepackage{color,dcolumn,graphicx,hyperref}
\hypersetup
{
  colorlinks = true, linkcolor = blue, citecolor = blue, urlcolor = blue,
}

@@

\begin{table}[ht]
\caption{Data sources used in taxize} % title of Table
\centering % used for centering table
\begin{tabular}{|l|ccc|c|}
\hline
Source name &  Name search & Name resolution & Phylogeny & URL  \\
\hline
Encyclopedia of Life & Yes & See GNR below & No & \url{http://eol.org/} \\
Integrated Taxonomic Resolution Service & Yes & Synonyms & No & X\\
iPlant Taxonomic Name Resolution Service & Yes & Yes & No & X \\
Phylomatic & No & X & No & X \\
uBio & Yes & X & No & X \\
Global Names Resolver (EOL) & Yes & X & No & X \\
\hline
\end{tabular}
\label{table:nonlin} % is used to refer this table in the text
\end{table}

@@

<<itisclass, eval=TRUE>>=
specieslist <- c("Abies procera","Pinus contorta")

classification(get_tsn(specieslist, "sciname"))
@

You can also get this type of information from the NCBI by doing \Sexpr{classification(get_uid(specieslist))}

@@

text here

@@

\subsection{Use ITIS locally for faster searches}

text

\subsection{Use ITIS locally for faster searches}

text

>>>>

\documentclass[letterpaper,superscriptaddress,showkeys,longbibliography]{revtex4-1}
\usepackage[utf8]{inputenc}
\usepackage{color,dcolumn,graphicx,hyperref}
\usepackage{natbib}
\bibliographystyle{ecol_let}
\hypersetup
{
  colorlinks = true, linkcolor = blue, citecolor = blue, urlcolor = blue,
}

@@

\begin{table}[ht]
\caption{Data sources used in taxize} % title of Table
\centering % used for centering table
\begin{tabular}{|l|ccc|c|}
\hline
Source name &  Name search & Name resolution & Phylogeny & URL  \\
\hline
Encyclopedia of Life & Yes & See GNR below & No & \url{http://eol.org/} \\
Integrated Taxonomic Resolution Service & Yes & Synonyms & No & X\\
iPlant Taxonomic Name Resolution Service & Yes & Yes & No & X \\
Phylomatic & No & X & No & X \\
uBio & Yes & X & No & X \\
Global Names Resolver & Yes & X & No & X \\
Global Names Index & Yes & X & No & X \\
IUCN Red List & Yes & X & No & X \\
Tropicos & Yes & X & No & X \\
Plantminer & Yes & X & No & X \\
Theplantlist.org & Yes & X & No & X \\
Catalogue of Life & Yes & X & No & X \\
National Center for Biotechnology Information (NCBI) & Yes & X & No & X \\
\hline
\end{tabular}
\label{table:nonlin} % is used to refer this table in the text
\end{table}

@@

<<itisclass, eval=TRUE>>=
specieslist <- c("Abies procera","Pinus contorta")
classification(get_tsn(specieslist, "sciname"))
@

It turns out both species are in the family Pinaceae. You can also get this type of information from the NCBI by doing \emph{classification(get\_uid(specieslist))}

Instead of a full classification, you may only wan a single name, say a family name for your species of interest. The function \emph{tax_name} is built just for this purpose. And you can specify the data source you retrieve the taxonomic name from with the \emph{db} parameter.

<<taxname, eval=TRUE>>=
tax_name(query="Helianthus annuus", get="family", db="itis")
tax_name(query="Helianthus annuus", get="family", db="ncbi")
@

@@

There is an increasingly common use case: many biologists are not adequately trained in reconstructing phylogenies. However, so-called \emph{taxonomic phylogenies} can be constructed from simply knowing the higher taxonomic classification of a set of taxa. There are few taxon groups for which we can get phylogenies simply based on taxonomy; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that lets you input a simple species list, and then taxize does work behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example.

@@

\subsection{What taxa are in children of my taxon of interest?}

If you aren't a taxonomic specialist on a particular taxon you likely don't know what children taxa are within a family, or within a genus. You can of course go to a website like Wikispecies (\url{http://species.wikimedia.org/wiki/Main_Page}) or Encyclopedia of Life (\url{http://eol.org/}). taxize provides an easy way for you to search for downstream taxa, both for the Catalogue of Life (CoL; \url{http://www.catalogueoflife.org/}) and the Integrated Taxanomic Information Database (\url{http://www.itis.gov/}). Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).

<<downstream, eval=TRUE>>=
col_downstream(name="Apis", downto="Species")[[1]]
@

\subsection{IUCN Status}

There are a number of things we can do once we have the correct taxonomic names. One thing we can do is ask about the conservation status of a species. We have provided a set of functions, \emph{iucn\_summary} and \emph{iucn\_status}, to search for species names, and extract the status information, respectively. Here, we search for the Panther and Lynx.

<<XXXX, eval=TRUE>>=
ia <- iucn_summary(c("Panthera uncia", "Lynx lynx"))
iucn_status(ia)
@

It turns out that the Panther is endangered (EN) and the Lynx is of least concern (LN).

### 2013-04-27 inst/doc/taxize_withcode.tex
updated the mansucript

\usepackage{alltt}
\usepackage[utf8]{inputenc}
\usepackage{color,dcolumn,graphicx,hyperref}
\hypersetup
{
  colorlinks = true, linkcolor = blue, citecolor = blue, urlcolor = blue,
}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
specieslist <- \hlfunctioncall{c}(\hlstring{"Abies procera"}, \hlstring{"Pinus contorta"})

\hlfunctioncall{classification}(\hlfunctioncall{get_tsn}(specieslist, \hlstring{"sciname"}))
\end{alltt}
\begin{verbatim}

@@

You can also get this type of information from the NCBI by doing list(ScientificName = c("cellular organisms", "Eukaryota", "Viridiplantae", "Streptophyta", "Streptophytina", "Embryophyta", "Tracheophyta", "Euphyllophyta", "Spermatophyta", "Coniferophyta", "Coniferopsida", "Coniferales", "Pinaceae", "Abies", "Abies procera"), Rank = c("no rank", "superkingdom", "kingdom", "phylum", "no rank", "no rank", "no rank", "no rank", "no rank", "no rank", "class", "order", "family", "genus", "species"), UID = c("131567", "2759", "33090", "35493", "131221", "3193", "58023", 
"78536", "58024", "3312", "58019", "3313", "3318", "3319", "97175")), list(ScientificName = c("cellular organisms", "Eukaryota", "Viridiplantae", "Streptophyta", "Streptophytina", "Embryophyta", "Tracheophyta", "Euphyllophyta", "Spermatophyta", "Coniferophyta", "Coniferopsida", "Coniferales", "Pinaceae", "Pinus", "Pinus", "Pinus contorta"), Rank = c("no rank", "superkingdom", "kingdom", "phylum", "no rank", "no rank", "no rank", "no rank", "no rank", "no rank", "class", "order", "family", "genus", "subgenus", "species"), UID = c("131567", "2759", "33090", "35493", 
"131221", "3193", "58023", "78536", "58024", "3312", "58019", "3313", "3318", "3337", "139271", "3339"))

@@

\subsection{Use ITIS locally for faster searches}

text

\subsection{Use ITIS locally for faster searches}

text

>>>>

\usepackage{alltt}
\usepackage[utf8]{inputenc}
\usepackage{color,dcolumn,graphicx,hyperref}
\usepackage{natbib}
\bibliographystyle{ecol_let}
\hypersetup
{
  colorlinks = true, linkcolor = blue, citecolor = blue, urlcolor = blue,
}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
specieslist <- \hlfunctioncall{c}(\hlstring{"Abies procera"}, \hlstring{"Pinus contorta"})
\hlfunctioncall{classification}(\hlfunctioncall{get_tsn}(specieslist, \hlstring{"sciname"}))
\end{alltt}
\begin{verbatim}

@@

It turns out both species are in the family Pinaceae. You can also get this type of information from the NCBI by doing \emph{classification(get\_uid(specieslist))}

Instead of a full classification, you may only wan a single name, say a family name for your species of interest. The function \emph{tax_name} is built just for this purpose. And you can specify the data source you retrieve the taxonomic name from with the \emph{db} parameter.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"itis"})
\end{alltt}
\begin{verbatim}

Retrieving data for species ' Helianthus annuus '
      family
1 Asteraceae
\end{verbatim}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"ncbi"})
\end{alltt}
\begin{verbatim}

Retrieving data for species ' Helianthus annuus '
      family
1 Asteraceae
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\subsection{What taxa are in children of my taxon of interest?}

If you aren't a taxonomic specialist on a particular taxon you likely don't know what children taxa are within a family, or within a genus. You can of course go to a website like Wikispecies (\url{http://species.wikimedia.org/wiki/Main_Page}) or Encyclopedia of Life (\url{http://eol.org/}). taxize provides an easy way for you to search for downstream taxa, both for the Catalogue of Life (CoL; \url{http://www.catalogueoflife.org/}) and the Integrated Taxanomic Information Database (\url{http://www.itis.gov/}). Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{col_downstream}(name = \hlstring{"Apis"}, downto = \hlstring{"Species"})[[1]]
\end{alltt}
\begin{verbatim}
  childtaxa_id     childtaxa_name childtaxa_rank
1      6971712 Apis andreniformis        Species
2      6971713        Apis cerana        Species
3      6971714       Apis dorsata        Species
4      6971715        Apis florea        Species
5      6971716 Apis koschevnikovi        Species
6      6845885     Apis mellifera        Species
7      6971717   Apis nigrocincta        Species
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection{IUCN Status}

There are a number of things we can do once we have the correct taxonomic names. One thing we can do is ask about the conservation status of a species. We have provided a set of functions, \emph{iucn\_summary} and \emph{iucn\_status}, to search for species names, and extract the status information, respectively. Here, we search for the Panther and Lynx.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
ia <- \hlfunctioncall{iucn_summary}(\hlfunctioncall{c}(\hlstring{"Panthera uncia"}, \hlstring{"Lynx lynx"}))
\hlfunctioncall{iucn_status}(ia)
\end{alltt}
\begin{verbatim}
Panthera uncia      Lynx lynx 
          "EN"           "LC" 
\end{verbatim}
\end{kframe}
\end{knitrout}

It turns out that the Panther is endangered (EN) and the Lynx is of least concern (LN).
